---
title: "simple simulation scenarios"
output: github_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Simplest simulation - TE invasion without host defence

## no host defence - mate radius 1

```{bash eval=FALSE}
# simulation code
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit none --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --file-spatial simple.txt 
```

**parameters explained**

* --seed 5: the random seed
* --genome MB:1,1: a genome of 2 chromosomes each with a length of 1MB; a TE insertion may occupy any site in this genome of 2MB; novel insertions will get random positions in the genome; if a site is already occupied the novel insertion will be ignroed
* --rr 4,4: a recombination rate of 4cM/Mb per chromosome
* --u 0.1: the transposition rate is u=0.1; the number of novel insertions per gamete is poisson distributed with lambda = cu/2, where c is the total number of insertions in the genome of the parent. why divided by two? because we have to assume that novel insertions may occur in any of the two genomes of a diploid
* --grid-x 50: the x-size of the 2D grid is 50
* --grid-y 20: the y-size of the 2D grid is 20
* --mate-radius 1: the mate radius is 1; i.e. when generating the next generation, every individual in the spatial grid is replaced by a new individual the potential mates yielding the two gametes for the novel individual will be found in a radius of 1 of the focal individual. e.g. if the focal individual has y-x coordinates 3-3, then the potential mates may have coordinates 2-2, 2-3, 2-4, 3-2, 3-3, 3-4, 4-2, 4-3, 4-4; note that the focal species could, by chance, also be one of the two mates
* --epi-inherit none: so far we assume no epigenetic inheritance of the TE silencing (more importantly so far we assume no host defense which would require providing the parameter `--trigger-threshold`, see below)
* --basepop 9-11,1-2,10: all individuals in the y-range 9-11 and the x-range 1-2 will have exactly 10 TE insertions; the insertions will be randomly distributed in the genome (of 2Mb). Hence every TE insertion will have a population frequency of 1/2N where N is the population size (N=`--grid-x` * `--grid-y`)
* --rep 1: we simulate a single replicate
* --steps 5: output will be generated all 5 generations; note that we perform additional filtering in R, hence the final figures in this tutorial do not show the results for all 20 time points for which we generated an output
--gen 100: simulations will be performed for 100 generations
--file-spatial simple.txt: the output file for generating the figure seen below

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs, c(seq(-1, 900, 300),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B") # "#db2c3fff",

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** the TE starts spreading at the small rectangular patch at generation 1 defined by `--basepop 9-11,1-2,10`; The TE is invading the population unconstrained by any host defense (therefore only green dots). Due to the absence of a host defense very high TE copy numbers per individual are reached by generation 100 (some have more than 900 insertions per diploid individual).

## no host defence - mate radius 2

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 2 --epi-inherit none --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --file-spatial simple-mr2.txt 
```

**novel parameters explained**

* --mate-radius 2: compared to the previous example we extend the mate radius to two, hence potential mates can now be found more distantly of the focal species. E.g. if the focal species has y-x coordinates of 3-3, than potential mates may for example have the coordinates 1-1, 1-5, 5-5, 3-3 etc

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-mr2.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs, c(seq(-1, 900, 300),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B") # "#db2c3fff",

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** with a larger mate radius the TE spreads faster


## no host defence - mate radius 'panmictic'

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius pan --epi-inherit none --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --file-spatial simple-pan.txt
```

**novel parameters explained**

* --mate-radius pan: panmictic simulations will be performed where the spatial coordinates are ignored. Hence everyone may mate with everyone; This can be quite useful. First we can test the simulator against theoretical population genetics models (that typically assume panmixis). Second we can evaluate the impact of spatial information on the invasion dynamics compared to panmixis.

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-pan.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,5,20,  50))

# bin
df$TEbin <- cut(df$TEs, c(seq(-1, 30, 5),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B") # "#db2c3fff",

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note**  Every individual still has spatial coordinates during the panmictic simulations. However, the coordinates are ignored for finding mates. Hence, the coordinates have **no** impact in the panmictic simulations. As a result the TE spreads quite stochastically in the spatial grid.

# Host defence 

## host defence - no epigenetic inherited silencing

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit none --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --trigger-defense 40 --file-spatial simple-defence.txt
```

**novel parameters explained**

* --trigger-defense 40: **simulations with host defense** the host defense will be triggered in any individual having 40 or more TE copies. In individuals with a host defense the TE will have a transposition rate of `--uc` wich is uc=0.0 by default.

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-defence.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs, c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** the TE starts spreading at the small patch by generation 1. By generation 50 already several indivdiuals have triggered the host defence (no epigenetic inheritance in this scenario); By generation 100, bot the TE and the host defence is spreading; **important** only individuals with 40 or more TE copies can have the silencing state

## host defence - uniparental epigenetic silencing inheritance

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit dros --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --trigger-defense 40 --file-spatial simple-defence-dros.txt 
```

**parameters explained**

* --epi-inherit dros: uniparental epigenetic inherited silencing; silencing is transmitted maternally as in Drosophila. since hermaphrodites are simulated one mate is randomly chosen as the mother; if a TE is silenced in the mother any offspring with at least one TE insertion will inherit the silencing-status. offspring that per chance do not have a single insertion will not inherit the silencing status; paternal offspring will have active TEs (unless the mother has silenced TEs)

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-defence-dros.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** in this scenario even individuals with <40 TEs may have silenced TEs (compare to previous scenario with --epi-inheritance none )

## host defence - biparental epigenetic silencing inheritance

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit ara --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --trigger-defense 40 --file-spatial simple-defence-ara.txt 
```

**parameters explained**

* --epi-inherit ara: biparental epigenetic inherited silencing; if a TE is silenced in any parent, the TE will be silenced in all offspring (even if the TE is still active in the other parent)

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-defence-ara.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** how the epigenetic silencing is spreading faster than the TE such that by generation 100 all TE insertions are silenced; furthermore 

## host defence - biparental epigenetic silencing and selfing 

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit ara --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --trigger-defense 40 --selfing-rate 0.95 --condinv --file-spatial simple-defence-ara-self.txt 
```

**parameters explained**

* --selfing-rate 0.95: there is a 95% chance that the mother and the father will be identical 
* --condinv: conditional on invasion; this is just a convenience method (rescuing me from finding seeds where it does work)

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-defence-ara-self.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

# Negative selection against inactive TEs 

## negative selection - biparental inheritance


```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit ara --basepop "1-20,1-50,40'" --rep 1 --steps 5 --gen 400 --trigger-defense 40 --s 0.1 --h 0.5 --file-spatial simple-negsel-ara.txt
```

**parameters explained**

* --s 0.1: selection against homozygous insertions, fitness w=1-s
* --h 0.5: selection against heterozygous insertions, fitness is w=1-hs
* --basepop "1-20,1-50,40'" all individuals in the entire population will have 40 **silenced** TE insertions (the dash indicates the silencing)
```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-negsel-ara.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  100,  200))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** selection against TEs may lead to a patchy distribution **when TEs are silenced biparental**

## negative selection - uniparental inheritance

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit dros --basepop "1-20,1-50,40'" --rep 1 --steps 5 --gen 400 --trigger-defense 40 --s 0.1 --h 0.5 --file-spatial simple-negsel-dros.txt 
```


```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-negsel-dros.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  100,  200,400))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Now this is a surprise to me! negative selection against TEs with uniparental silencing does not lead to patchyness** actually it reactivates the TE on a broad scale, even when the TE was silenced in the entire base population! thats truly fascinating! but makes sense, as soon as some individuals  have zero TEs (thanks to neg.sel.) the TE can be reactivated by paternal crosses


# Population structure - uniparental inheritance

```{bash eval=FALSE}
./invade --seed 5 --genome MB:1,1 --rr 4,4 --u 0.1 --grid-x 50 --grid-y 20 --mate-radius 1 --epi-inherit dros --basepop 9-11,1-2,10 --rep 1 --steps 5 --gen 100 --trigger-defense 40 --mate-barrier 25 --barrier-strength 0.99 --file-spatial simple-matebar.txt 
```

**parameters explained**

* --mate-barrier 25: position of mating barrier in the x-grid; it will be more difficult for individuals at opposite ends of the barrier to mate eg with x-coordinates 25 and 26
* --barrier-strength 0.99: a 99% reduction in mating-probabilty for individuals on opposite ends of the barrier

```{r echo=FALSE}
library(tidyverse)
df <- read_tsv("/Users/robertkofler/analysis/2025-scenarious/simple-matebar.txt") %>% select(-1)

# filter generations
df <-  df %>% filter(gen %in% c(0,  50,  100))

# bin
df$TEbin <- cut(df$TEs,  c(seq(-1, 40, 10),Inf))

# assign colors
bin_colors <- c("darkgrey", "#3DA53B","#db2c3fff") 

# generate plot
pse <- ggplot(df, aes(x = X, y = Y)) +
  geom_point(aes(alpha = TEbin, color = silenced), size = 2) +
  scale_color_manual(values = bin_colors) +  # Use discrete bins
  scale_alpha_ordinal(range=c(0.5,1.0))+
  facet_grid(rows = vars(gen), switch = "y") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
       axis.title = element_blank(),
        strip.text.y = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.title = element_text(size = 14),  # Legend title size
       legend.text = element_text(size = 12)) +
  labs(color = "TE Status", alpha = "Copynumber")
# save to file
plot(pse)

```

**Note** the coordinates of the mate barrier can be clearly discerned at generation 100; also not how all individuals in the left subpopulation have about 40 TE copies (ie the silencing threshold), whereas most individuals in the right subpopulation have very few insertions